Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-18 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
fix related to #110

closes #109

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]. 
The latter can result from a combination of still air at lower heights (Fig. `r fig_NEON_vertical`e), and sunflecks or the 'canopy greenhouse effect', as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [Fig. `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. 
One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018]. 
However, the opposite trend was observed for a lone tree surrounded by grass in the same area, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].

@@

Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
In addition, shading by the canopy can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020]. 
As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based Forward Looking Infrared camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD. 
However, the authors did not consider vertical light environment conditions even though light directly influences the deviation between canopy temperature and $T_{air}$. 
In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig. `r SIfig_NEON_all`).
The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.

>>>>

Vertical $T_{leaf}$ gradients are also expected to vary with canopy structure (Fig. `r fig_leaf_T`). 
Forests with closed canopies and high LAI, including tropical and temperate broadleaf forests, act as a parasol, absorbing most of the incoming radiation and preventing vertical air mixing in the understory. 
Therefore, in these forests, leaves in the upper canopy can experience greater $T_{leaf}-T_{air}$ and higher maximum $T_{leaf}$ than do lower-canopy leaves, in some cases exceeding the optima for photosynthesis ($T_{opt}$) because of $g_s$ limitation and high solar radiation loads [@carterExperimentalWarmingTropical2021; @niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, open canopies with lower LAI allow more vertical air mixing and sunlight into the understory. 
This mixing and light transmission can either neutralize a $T_{leaf}$ gradient or elevate $T_{leaf}-T_{air}$ at lower heights relative to the upper canopy [Fig. `r fig_NEON_vertical`h, Supporting Information Figure `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002]. 
The latter can result from a combination of still air at lower heights (Fig. `r fig_NEON_vertical`e), and sunflecks or the 'canopy greenhouse effect', as mentioned above [@hardwickRelationshipLeafArea2015; @schymanskiStomatalControlLeaf2013], and because shade leaves tend to have lower $g_s$  [@schymanskiStomatalControlLeaf2013]. 
Similarly, in very open forests or savannas, trees growing close to the ground can experience greater heat stress in their lower than upper canopies due to heat from the soil increasing the $T_{leaf}$ of the foliage close to the ground [Fig. `r SIfig_NEON_all`, @hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019; @johnstonWhatLiesVertical2020]. 
One of the few remote sensing studies combining drone lidar and thermal data found strong vertical gradients in midday plant temperature with ~5$^{\circ}$C cooler temperatures lower in the vertical profile of a dense forest stand in the sub-alpine Eastern Swiss Alps [@websterThreedimensionalThermalCharacterization2018]. 
However, the opposite trend was observed for a lone tree surrounded by grass in a California open oak woodland, with cooler temperatures at the top of the tree crown, indicating a strong influence of closed-canopy shading on vertical temperature gradients [@johnstonWhatLiesVertical2020].

@@

Detailed structural information from lidar data has been used to measure tree height, DBH, and crown dimensions [@fisherECOSTRESSNASANext2020], evaluate spatial variation in vertical leaf area density profiles [@dettoSpatialVariabilityTropical2015], and resolve complex seasonal and diurnal variation in shortwave radiation forcing in ecosystems [@musselmanEstimationSolarDirect2013].
In addition, shading by the canopy can be better represented to predict understory temperatures by modeling the time-varying interaction of the forest canopy with solar radiation, as has been done in other lidar-based analyses [e.g., @davisLiDARderivedTopographyForest2019].
Satellite and airborne thermal infrared remote sensing analyses are now being conducted at regional and continental scales, although very few applications exist at finer-scales necessary for understanding of vertical variation in plant canopy temperatures [@johnstonWhatLiesVertical2020]. 
As an example of the type of insight that can be gained from this approach, @pauTropicalForestTemperature2018 used data from a tower-based Forward Looking Infrared camera in combination with eddy-covariance data and found that tropical forest canopy temperatures were more strongly associated with GPP than $T_{air}$ or VPD.
In a savanna system in the western U.S., @johnstonWhatLiesVertical2020 found significant vertical variation in plant temperature, with warmer temperatures in the understory compared to the top of the blue oak canopies -- a finding that occurred in association with very high grass temperatures and consistent with the principles outlined in sections 1-2 (Fig. `r SIfig_NEON_all`).
The growing availability of these data makes this an opportune time to link fine-scale and landscape-scale measurements with the type of *in situ* measurements reviewed above to further explore ecosystem-scale patterns in vertical temperature gradients seasonally and across biomes.
